---
title: "R Workshop UCSD"
author: "Ben Wilson"
date: "March 3, 2016"
output: 
  revealjs::revealjs_presentation:
    theme: black
    reveal_options:
      slideNumber: true
      previewLinks: true
    highlight: pygments
    stylesheet: custom.css
---

```{r setup, include=FALSE}
knitr::dep_prev()
knitr::opts_chunk$set(
  echo=TRUE,
  tidy=FALSE,
  highlight=FALSE,
  cache=TRUE,
  messages=FALSE,
  warnings=FALSE,
  fig.height=4
  )

r <- getOption("repos")
  r["CRAN"] <- "http://cran.cnr.berkeley.edu/"
  options(repos = r)
```

# R 101

## Content Overview

  - R & Rstudio
  - Base R
  - Reading Data
  - Packages
  - Machine Learning

# Why R?

## Why R?

  - http://pypl.github.io/PYPL.html
    - Ranked 10 in popularity (Mar 2016)  

  - http://blog.revolutionanalytics.com/2015/11/new-surveys-show-continued-popularity-of-r.html
    - Skill second only to data analytics (roughly tied with python)


## What is R?

R is a language and environment for statistical computing and graphics.

- Compared to python
    - Built for data analysis instead of a general programming language
    - Python + Pandas are a similar tool kit
    - Easily cross platform
    
- Compared to Matlab
    - Built for 'data.frames' vs matrices
    - More flexible with data types
    - Free
    
    
## When Should I Use R?

R is useful for investigating data, creating models, and reports among other things.

Personally I use R to investigate new data, prototype models, and do ad-hoc analysis.

http://shiny.rstudio.com/gallery/


## WCGW

What could go wrong?

  - In memory calculations
  

# Base R

## Base R

```{r base_R}
2 + 3

8 * 6

8 * 6 == 5

cos(pi)
```


## Data Frames

```{r data_frames}
n <- c(2, 3, 5) 
s <- c("aa", "bb", "cc") 
b <- c(TRUE, FALSE, TRUE) 
df <- data.frame(n, s, b)

df
```

```{r column_addition}
df$runif <- runif(3)
df
```

## Data Frames (cont)

```{r str}
str(df)
```

```{r summary}
summary(df)
```

## WCGW

Type conversions. One common issue is types are incorrect for a given application.

```{r as.numeric}
df$n <- as.numeric(df$n)
df$s <- as.character(df$s)
str(df)
```

```{r apropos}
?mean
apropos('mean')
```


# Reading Data into R

## Reading Data into R

Data provided by https://archive.ics.uci.edu/ml/datasets/Online+News+Popularity. Goal is to predict number of shares from metadata of a link on mashable.com.


```{r read_csv}
df <- read.csv('data/OnlineNewsPopularity.csv')
nrow(df)
```


K. Fernandes, P. Vinagre and P. Cortez. A Proactive Intelligent Decision Support System for Predicting the Popularity of Online News. Proceedings of the 17th EPIA 2015 - Portuguese Conference on Artificial Intelligence, September, Coimbra, Portugal.


## Names

```{r names_df}
names(df)
```


## Investigate Data

```{r summary_df}
summary(df)
```


## Single Column Investigation

```{r shares}
head(df$shares)
mean(df$shares)
max(df$shares)
```

## Single Column Investigation
```{r hist}
hist(df$title_sentiment_polarity)
```


## Correlation Investigation
```{r corr}
plot(df[1:500, c("num_hrefs", "title_sentiment_polarity", "is_weekend", "shares")])
```

## WCGW

Delimited files are the most commin file ingestion type. A generic reader is read.table.

```{r read_table, eval=FALSE}
tab_df <- read.table(
  'my_files.tab', 
  sep='\t', 
  header=FALSE, 
  quote="'"
)
```

```{r names, eval=FALSE}
names(tab_df) <- c('foo', 'bar', 'baz')
```

# Functions and Logic

## Functions

```{r function_example, error=TRUE}
square <- function(x){
  return(x^2)
}

square(5)

square('foo')

square(NA)
```

## Logic

```{r logic}
if (5 > 10){
  print('baz')
}
```


```{r vectorized}
unif_sample <- runif(100)

unif_test <- ifelse(unif_sample > .50, 'high', 'low')

table(unif_test)
```


## WCGW

In real data, NAs and NULLS are pervasive. Many functions handle these properly, e.g. max(5, NA) = NA. 

```{r na}
sum(unif_sample, na.rm=TRUE)
```

Writing for loops is possible, but slow. Avoid when possible
```{r loop}
for (i in 1:length(unif_sample)) {
  unif_test[i] <- unif_sample[i] > .50 
}
```

# Packages

## Packages in R

R has a built in package manager. Packages (and R source code, binaries) are hosted on a distributed network (CRAN). http://cran.us.r-project.org/ . There are currently more than 8000 packages available + many more on github. Packages can be written in R, C++, Java and many more languages.

To install packages one uses the command install.packages() in the R terminal.

## Install a couple of packages

Two packages which enable data manipulation and visualization are dplyr and ggplot2. To install them simply type

```{r install, eval=FALSE}
install.packages(c('dplyr', 'ggplot2'))
```

```{r load_package, results=FALSE}
library(dplyr)
library(ggplot2)
```

## Dplyr

Dplyr introduces (at least) two ideas to data manipulation in R.

  - Pipes
```{r pipes}
  df$shares %>% max()
```  
  - Verbs
```{r verbs}
  df %>% summarize(n=n())
```


## Group By

Dplyr also introduces many SQL-like verbs in an accessible fashion.

```{r group_by}
df %>% 
  group_by(is_weekend) %>% 
  summarize(
    n=n(),
    shares_mean=mean(shares),
    shares_max=max(shares)
  )
```

## WCGW

Sometimes a package isn't what you want.

```{r undo, eval=FALSE}
detach("package:idontwantthis", unload=TRUE)
remove.packages('idontwantthiseither')
```


# GGPlot2

## GGPlot2

ggplot2 is a plotting package which enables extremely customizable graphs in an intuitive way.

```{r ggplot, fig.height=3}
ggplot() +
  geom_histogram(
    data=df,
    aes(x=title_sentiment_polarity)
  )
```


## WCGW

Greater than 8000 packages to choose. Lots of freedom and subtle differences. 

Vignettes provided a lot of information for a package.

Ex: https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html

Ex: http://docs.ggplot2.org/current/

Reading r-bloggers + stack-overflow helps as well.


# Machine Learning

## Machine Learning Packages

There are many packages devoted to machine learning in R. Some packages for those interested

  - caret
  - e1071 (a collection of methods)
  - nnet
  - gbm
  - RWeka
  - party


## Sampling Data

Splitting the data into testing and training sets is essential.

```{r data_split}
set.seed(290875)

train_rows <- 
  sample(nrow(df), nrow(df) * .10)

train_df <-
  df %>%
  slice(train_rows)

test_df <-
  df %>%
  slice(-train_rows)

head(train_rows)
```


## Running a random forest

```{r party_install, eval=FALSE}
install.packages('party')
library(party)
```


```{r party, echo=FALSE, results='hide'}
library(party)
```

```{r rf}
df_cforest <- 
  cforest(
    shares ~ avg_positive_polarity + avg_negative_polarity + global_subjectivity + global_sentiment_polarity + global_rate_positive_words + global_rate_negative_words, 
    data=train_df,
    control = cforest_unbiased(mtry = 2, ntree = 50)
  )
```

## Predicting New Values

```{r predict}
test_df <-
  test_df[1:1000, ]

test_df$prediction <-
  Predict(df_cforest, newdata=test_df)

sqrt(mean((test_df$shares - test_df$prediction)^2))
mean(abs(test_df$shares - test_df$prediction))
```

## Plotting The Error

```{r warning_trick, echo=FALSE, results='hide'}
options(warn=-1)
```


```{r predict_plot_code, warnings=FALSE, results='hide', fig.show='hide'}
test_df$is_weekend <- as.factor(test_df$is_weekend)

ggplot() +
  geom_density(
    data=test_df,
    aes(
      x=shares-prediction, 
      fill=is_weekend, 
      group=is_weekend, 
      color=is_weekend
    ),
    alpha=.4
  ) +
  xlim(-5000, 5000) +
  theme_bw() + 
  ggtitle("Shares - Prediction ")
```


## Plotting the Error
```{r predict_plot, echo=FALSE, warnings=FALSE, fig.height=4}
test_df$is_weekend <- as.factor(test_df$is_weekend)
ggplot() +
  geom_density(
    data=test_df,
    aes(x=shares-prediction, fill=is_weekend, group=is_weekend, color=is_weekend),
    alpha=.4
  ) +
  xlim(-5000, 5000) +
  theme_bw() + 
  ggtitle("Shares - Prediction")
```


## WCGW

  - In memory calculations
  
  - Column typing
  
  - Curse of dimensionality 
  
  